Determining the direction of causality in the face of measurement error

### 2016-02-25 docs/manuscript.Rmd
updated analysis

fig_nums(
	"cit_mr_comparison_figure", 
	caption="Outcomes were simulated to be causally influenced by exposures, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The value of $d = \\rho_{x, x_O} - \\rho_{x,y}\\rho_{y,y_O}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations.",
	display=FALSE
)

@@

With the explosion in data that measures the human phenome, of great interest is understanding the causal nature of the emerging putitive correlations between measures of interest. Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling. The principle behind mediation analysis can be explained as follows. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the causal inference test (CIT) [@Millstein2009], which have been employed by a number of recent publications make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

While mediation based approaches are simple and convenient to implement, it is important to note where they may lead to unreliable results. One such mechanism is measurement error [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013]. Measurement (or observational) error is the difference between a measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect sensitivity, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Sensitivity is also an issue, for example if the unit of measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error can indeed arise in more low-tech data too, for example when measuring body mass index (BMI) one is typically interested in using this as a proxy for obesity, but it is clear that the correlation between BMI and obesity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically have both bias and imprecision compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].

An alternative statistical method that gained prominence at least in part as a solution to the problem of measurement error [@Ashenfelter1994] is Mendelian randomisation (MR) [@DaveySmith2004; @DaveySmithHemani2014]. Here an instrumental variable (typically a SNP) that has a robost, causal association with the exposure is used to proxy the exposure in an association test with the outcome. If there is no measurement error in the SNP then the causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure. This approach also has the useful property that it guards against unmeasured confounders driving the association between exposure and outcome; and if the biological effect of the SNP is understood then it can also aid against reverse causality driving the association.

Here, however, arises a potential issue. Often the biological effect of a SNP is not known, and therefore it can be difficult to determine for which of the two variables in a putative exposure-outcome association is the SNP a valid instrument. By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. For example genome wide association studies (GWASs) that identify genetic associations for complex traits are by design hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

@@

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$. To infer the direction of causality using the CIT method, the omnibus p-value was estimated for $x$ causing $y$, and for $y$ causing $x$, and the model that returned the lowest p-value was taken as the one denoting the correct causal direction.

@@

and a p-value is generated from the $Z$ value to indicate confidence of difference in correlations $\rho_{gx}$ and $\rho_{gy}$. The test for a causal association is then obtained by performing two sample MR in the direction that is predicted to be the correct model based on the Z score test. For simplicity, the test is deemed to have obtained a robust association if the Z score test has a p-value $<  0.05$ and the two sample MR analysis has a p-value $< 0.05$. Note that the same approach can be applied to a two-sample MR setting, where

@@

We performed simulations to explore the performance of this approach in comparison to CIT. `r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, in all cases the MR analysis infers the wrong direction of causality at the same or a lower rate than the CIT. When $d > 0$ is satisfied we observe that in many cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.

>>>>

fig_nums(
	"cit_mr_comparison_figure", 
	caption="Outcomes were simulated to be causally influenced by exposures, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The value of $d = \\rho_{x, x_O} - \\rho_{x,y}\\rho_{y,y_O}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations. For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality.",
	display=FALSE
)

@@

With the explosion in data that assess the human phenome, we are presented with the challenge of understanding the causal nature of the emerging putitive correlations between measures of interest. Mediation is a commonly used statistical method for making causal inference in observational data [@Millstein2009; @Schadt2005; @Gayan2010a; @Tavtigian2009], and it exists in many forms, from simple regression based systems to structural equation modeling (SEM) [@Aten2008]. The principle behind mediation analysis can be explained as follows. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods, which have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

While mediation based approaches are simple and convenient to implement, it is important to note where they may lead to unreliable results. One such mechanism is measurement error [@LeCessie2012; @Nagarajan2013; @Shpitser2010; @Blakely2013], which has been noted to be an explicit issue with CIT style approaches [@Wang2015; @Lagani2015]. Measurement (or observational) error is the difference between a measured value of a quantity and its true value. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but they come with the problem of having imperfect sensitivity, so for example methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value [@Harper2013; @Chen2013a]. Sensitivity is also an issue, for example if the unit of measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more low-tech data too, for example when measuring body mass index (BMI) one is typically interested in using this as a proxy for obesity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically have both bias and imprecision compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the context of mediation analysis in the epidemiological literature extensively [@LeCessie2012; @Blakely2013].

An alternative statistical method that gained prominence as a solution to the problem of measurement error is Mendelian randomisation (MR) [@DaveySmith2004; @DaveySmithHemani2014]. Here an instrumental variable (typically a SNP) that has a robost, causal association with the exposure is used to stand for the exposure in an association test with the outcome. If there is no measurement error in the SNP then the causal association between the exposure and the outcome can be estimated by scaling the association between the SNP and the outcome by the association between the SNP and the exposure. Indeed this approach was in part pioneered (in the form of instrumental variable (IV) analysis) as a correction for measurement error [@Ashenfelter1994], but it also has the useful property that it guards against unmeasured confounders driving the association between exposure and outcome; and if the biological effect of the SNP is understood then it can also aid against reverse causality driving the association.

Here, however, arises a potential issue: often the biological effect of a SNP is not known. If there exists a putative association between two variables, with a SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the direct effect of the SNP (i.e., for which of the two variables is the SNP a valid instrument?). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, so if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. For example genome wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. Another situation is where the causal direction between 'omic measures need to be determined, for example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Waszak2015]?

@@

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The *cit.cp* function was used to obtain an omnibus p-value of causality between an exposure $x$ and an outcome $y$ using a mediator $g$ by making inferences based on a number of conditions that we examine in the results. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the correct model of $x$ causing $y$, and for incorrect model of $y$ causing $x$. For some significance threshold $\alpha$, causal direction was inferred based on the following scenarios:

1. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} > \alpha$ then the correct model is accepted
2. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} < \alpha$ then the incorrect model is accepted
3. If $p_{CIT,correct} > \alpha$ and $p_{CIT,incorrect} > \alpha$ then the an alternative causal model is accepted
4. If $p_{CIT,correct} < \alpha$ and $p_{CIT,incorrect} < \alpha$ then no inference is made

For the purposes of compiling simulation results we use an arbitrary $\alpha = 0.05$ value.

@@

and a p-value is generated from the $Z$ value to indicate confidence of difference in correlations $\rho_{gx}$ and $\rho_{gy}$. The test for a causal association is then obtained by performing two sample MR in the direction that is predicted to be the correct model based on the Z score test. For the purposes of compiling simulation results, the test is deemed to have obtained a robust association if the Z score test has a p-value $< 0.05$ and the two sample MR analysis has a p-value $< 0.05$, though in practice we do not advocate using an arbitrary $p$ value threshold for denoting significance in making causal inference.

Note that the same correlation test approach can be applied to a two-sample MR [@Pierce2013] setting, where the Steiger test of two independent correlations is applied instead of the case detailed here where correlations share one variable in common. An advantage of using the Steiger test in the two sample context is that it can compare correlations in independent samples where sample sizes are different.

@@

We performed simulations to explore the performance of this approach in comparison to CIT. `r cfig("cit_mr_comparison_figure")` shows that, as predicted, when $d < 0$ the MR analysis is liable to infer the wrong direction of causality, and that this erroneous result is more likely to occur with increasing sample size. However, in a large number of cases the MR analysis infers the wrong direction of causality at a much lower rate than the CIT. When $d > 0$ is satisfied we observe that in most cases the MR method has greater power to obtain evidence for causality than CIT, and always obtains the correct direction of causality.
